Numerical trajectory patterns of floating macroalgae

### 2021-01-10 thesis_chapter_3.Rmd
chapter 3 discussion updated

The general flow in the study domain is north-westward and is characterised by sub-mesoscale processes such as vortices and sub-mesoscale eddies.

Past research has inferred from direct and indirect techniques that drifting macroalgae tend to follow general flow patterns in the region in question. For example...

The results from this numerical experiment confirm that drifting macroalgae do follow the surface current in the study region when considering only hydrodynamic drag. In addition, not only are the trajectories similar but so are the end points at the end of simulation. However, when ad-hoc wind drag (windage) is considered in combination with hydrodynamic drag the trajectories differ greatly when compared to purely langrangian particle trajectories. The inclusion of wind drag causes the particles to flow further away from the coast and with a greater distance travelled from the release location. The difference in trajectory and distance covered can be attributed to the inclusion of wind drag. Wind drag adds to the effects of hydrodynamic drag when the wind direction and current direction are opposite to each other. However, when the wind direction is in the same or similar direction to that of the current, momentum energy from the wind is transferred to the object. This causes the particles to travel further and to be exposed to different time varying flows. insert putman study here. This is also reflected in the analysis comparing distances travelled by all the particles in each simulation. The particles in the simulations that included wind drag travlled significantly greater distances compared to the passive or hydrodynamic only drag.

The addition of drag forces into the simulation also causes the particles to cluster together along the mean trajectory compared to that of the passive particles which are more evenly dispersed. This is most likely due to the shared drag characteristics among the particles which included the plants weight and surface area into the calculation. Other studies investigating flotsam trajectory characteristics have found similar results. A study by @miron2020 conducted field experiments using a range of objects such as spheres, cubes, cuboids to compare to investigate the effect of inertia on particles dynamics. The results from the study showed that objects tended to cluster according to the shape of the object which the authors attributed to shared characteristics of inertia. Although drag forces were used in this study and not inertia, this study provides evidence to account for the higher density of particles along the mean trajectory.

@@

The same is also true when comparing different hydrodynamic and wind drag exposure scenarios. These scenarios reflect different buoyancy situations which result in different amounts of surface area exposed to hydrodynamic and/or wind drag. The similarities between the different combined drag scenarios suggests that higher magnitudes of wind exposure do not significantly alter trajectories. Instead, the results from this study suggest that the inclusion of drag forces in simulating macroalgal trajectory may result in improving accuracy of entrainment patterns. All the simulations that included any form of drag resulted in a fairly high density of particles entrained in a vortex. The entrainment and expulsion from eddies and vortices is an important characteristic of floating macroalgae. Accuracy in predicting entrainment and expulsion of macroalgae-rafts from eddies has been investigated previously by @putman2018 and @putman2020, who show that including windage greatly improves entrainment patterns when comparing virtual and real sargassum rafts. However, the aforementioned studies investigated large rafts and not floating individuals. Inertia is based on the size/weight of the raft, while hydrodynamic and wind drag are based on current velocity and surface area. Therefore, the role of inertia may be negligible for a solitary individual and rather other forms of drag should be considered. The results from this study suggest that the inclusion of hydrodynamic and/or wind drag is an important component when simulating virtual kelp particles will greatly increase the accuracy of entrainment and expulsion from sub-grid mesoscale oceanographic features, such as eddies and vortices.

??the study by putman which used different ad-hoc windage paramterisations showed that higher windsage resulted in greater accuracy of entrainment in eddies??.

>>>>

Past research has inferred from direct and indirect techniques that drifting macroalgae tend to follow the prevailing surface currents [@thiel2005, @thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. The results from this numerical experiment confirm that drifting macroalgae do follow the surface current in the study region when considering only hydrodynamic drag. In addition, not only are the trajectories similar but so are the end points at the end of simulation. However, when ad-hoc wind drag (windage) is considered in combination with hydrodynamic drag the trajectories differ greatly when compared to purely lagrangian particle trajectories. The inclusion of wind drag causes the particles to flow further away from the coast and with a greater distance travelled from the release location. The difference in trajectory and distance covered can be attributed to the inclusion of wind drag. Wind drag adds to the effects of hydrodynamic drag when the wind direction and current direction are opposite to each other. However, when the wind direction is in the same or similar direction to that of the current, momentum energy from the wind is transferred to the object. This causes the particles to travel further as well as exposed to different time varying flows. Past studies have shown this when including windage as a function of inertia [@putman2018;@brooks2019; @putman2020]. A study by @putman2020 used ad-hoc windage factors that was based on the surface current velocity to assess the approach for improving transport predictions of pelagic Sargassum. The results showed that including a ad-hoc windage factor in proving virtual trajectories to tracked Sargassum mats which was partly due to the inclusion of momemtum energy transfer from the wind. This is also reflected in the analysis comparing distances travelled by all the particles in each simulation. The particles in the simulations that included wind drag traveled significantly greater distances compared to the passive or hydrodynamic only drag.

The addition of drag forces into the simulation also causes the particles to cluster together along the mean trajectory compared to that of the passive particles which are more evenly dispersed. This is most likely due to the shared drag characteristics among the particles which included the plants weight and surface area into the calculation. Other studies investigating flotsam trajectory characteristics have found similar results [@breivik2011; @olascoaga2020; @miron2020]. A study by @miron2020 conducted field experiments using a range of objects such as spheres, cubes, cuboids to compare to investigate the effect of inertia on particles dynamics. The results from the study showed that objects tended to cluster according to the shape of the object which the authors attributed to shared characteristics of inertia. Although drag forces were used in this study and not inertia, the same conclusion applies, drag is an important characteristic to consider when investigating or predicting solitary drifting macrolagae.

@@

The same is also true when comparing different hydrodynamic and wind drag exposure scenarios. These scenarios reflect different buoyancy situations which result in different amounts of surface area exposed to hydrodynamic and/or wind drag. The similarities between the different combined drag scenarios suggests that higher magnitudes of wind exposure do not significantly alter trajectories. Instead, the results from this study suggest that the inclusion of drag forces in simulating macroalgal trajectory may result in improving accuracy of entrainment patterns. All the simulations that included any form of drag resulted in a fairly high density of particles entrained in a vortex. The entrainment and expulsion from eddies and vortices is an important characteristic of floating macroalgae. Accuracy in predicting entrainment and expulsion of macroalgae-rafts from eddies has been investigated previously by @putman2018 and @putman2020, who show that including windage greatly improves entrainment patterns when comparing virtual and real Sargassum rafts. However, the aforementioned studies investigated large rafts and not floating individuals. Inertia is based on the size/weight of the raft, while hydrodynamic and wind drag are based on current velocity and surface area. Therefore, the role of inertia may be negligible for a solitary individual and rather other forms of drag should be considered. The results from this study suggest that the inclusion of hydrodynamic and/or wind drag is an important component when simulating virtual kelp particles will greatly increase the accuracy of entrainment and expulsion from sub-grid mesoscale oceanographic features, such as eddies and vortices.

The use of a cylindrical shape coefficient instead of a sphere in the calculation of drag force also had no effect on the mean trajectory. Instead, the inclusion of a differnt shape coefficient increased to the density of particles entrained in an eddy field. This suggest that inclusion of different shape coefficients play a role in further improving estmiates and hindcasts of macroalgal disperal patterns.

# Conclusion

Most of the past research has focused on large macroalgae rafts which can vary greatly in size and shape, however no research exists attempting to clarify the drifting characteristics of solitary macroalgae. Understanding how to accurately model the distribution of solitary kelp can lead to be understanding of accumulation zones and sinks and ultimately the ecological links pertaining to those processes. The findings from this study indicate that solitary floating virtual *E. maxima* particles do tend to follow the prevailing surface currents but that the inclusion of wind causes the trajectories to differ greatly from that of a purely lagrangian particle. Furthermore, the inclusion of both hydrodynamic and wind drag cause clustering of particles along the trajectory as well causing particles to become entrained in eddies. Different cross-sectional areas exposed to hydrodynamic and wind drag had no effect on overall trajectory suggesting that those differences are neglible when investigating dispersal patterns of drifting *E.maxima*. In addition, the use of a cylindrical shape coefficient has no effect on mean trajectory but also causes higher density of particles to become entrained in an eddy. Overall the inclusion of drag forces is an important aspect to consider when investigating the dispersal patterns of solitary macroalgae, while differences in cross-sectional areas are not. This study also provides an approach which can be adapted to model floating solitary macroalgae, provided the surface area can be estimated accurately. However, more gaps in the research exist when considering what oceanographic processes play a role in solitary macroalgal dispersal as well as how these change seasonally. The identification of biological and physical factors that play a role in accumulation zones is also needed, as well as better estimates for sinking rates and raft-times.
